### 2021-04-25 assertions/doc/thesis-bernhardt/chapters/5-offline_design_tezos.tex
add missing orchestration decision
Revision compiler

\subsubsection{High-level languages}
Michelson is a compilation target for various high-level languages that provide a more user-friendly and intuitive way of writing Smart Contracts. Additionally, some of them come with development environments, testing or verification tools. The following list introduces three prominent languages compiling to Michelson:
\begin{description}
\item[\textbf{Liquidity}] \cite{liquidity} is a language with an OCaml-like syntax, which allows using local variables instead of stack manipulations. Its module system can be used to write reusable contract code or libraries. Besides an optimizing compiler, the project also includes a decompiler to compile Michelson programs to Liquidity source code.
\item[\textbf{SmartPy}] \cite{smartpy} is a language available through a Python library and lets developers write contracts and tests using Python syntax and structures (such as classes). Its developer suite includes, i.a., a compiler, a simulation engine for testing contracts and an online editor.
\item [\textbf{Ligo}] \cite{ligo} offers multiple syntax flavours close to Pascal, OCaml or Reason. Besides compilation, the toolchain allows to invoke or evaluate the generated code locally. The project provides Ligo as OCaml packages, which can be integrated into other projects.
\end{description}

@@

@@

\subsection{Compiler}\label{sec:compiler}
The transformed and type checked assertion AST needs to be retargeted to Michelson code according to the orchestration scheme. Since the modular approach is less complex to implement and avoids having to decompile the parent code, this section describes a compiler that assembles the assertion and contract code according to \figref{fig:modular_assembly}. The target language of the compiler of the pipeline can either be Michelson directly, or another high-level language which supports Michelson as a target. \secref{sec:tezos} lists some examples of such languages. The chosen language then serves as an intermediate representation (IR) of the assertion and manager code, which is ultimately compiled to Michelson by the compiler of the language. After specifying the steps of the compilation in a general way, the compilation approaches are described and discussed in more detail in \secref{sec:direct} and \ref{sec:IR}.

\subsubsection{Compilation of the assertion contracts}
Since the assertions are independent of each other and the manager code, each of them can be compiled in isolation to a separate Michelson program. The generated code of the assertion body and the parameter type can be plugged into a template that already contains the necessary boilerplate code for a contract.
% type inference from patterns
% handling of patterns like cons

\subsubsection{Compilation of the manager contract}
Given that the manager contract mirrors the parent contract and assuming that the proposal for the storage type from \lstref{lst:manager_modular} is adopted, the type declarations can be translated directly to the target language. Algorithm \ref{alg:compile_manager} shows the outline of a simplified algorithm that builds the code body recursively. For simplicity, entrypoints that take a sum type as parameter (e.g. \texttt{(or \%AB int int)} are not considered in the algorithm. However, in the actual compiler implementation they need to be taken care of. 
\begin{algorithm}
\caption{Simplified recursive algorithm for building the manager contract}\label{alg:compile_manager}
	\begin{algorithmic}[0]
	\State global storage = \{0: <parent>\} \Comment{Stores the mapping of ids to entrypoints}
	\State global i = 1 \Comment{Counter generating unique ids}
	\Function{compile}{parent parameter type}
	\If{ty = Or(l,r)}
	\State code\_left = compile(l)
	\State code\_right = compile(r)
	\State \Return code\_left + code\_right
	\Else
	\If{assertion exists for this entrypoint}
	\State code = generate\_loop(ty, i)
	\State storage.add(i++, entrypoint tag)
	\Else
	\State code = generate\_forward(ty)
	\EndIf
	\State \Return code
	\EndIf
	\EndFunction
	\end{algorithmic}
\end{algorithm}

The function \texttt{generate\_loop} generates code that does the following:
\begin{enumerate}
\item Retrieve the assertion contract and parent contract addresses from storage
\item Calculate $t$
\item Build a list of $t$ internal operations that call the assertion contract
\item Return the list of emitted operations and the unchanged storage
\end{enumerate}
% calculate size of search space? -> need to derive a formula from range constraints
By passing the next id $i$ to the function, the access to the storage can be hard-wired into the generated code. The address of the parent contract is assumed to always be stored with the id 0. \secref{sec:prob_threshold} stated a formula for calculating $t_{total}$ with the parameters $n$, the size of the search space, and the probability threshold $c$. While $n$ can be calculated from the ranges of the random generator(s), $c$ must be given as parameter to the compiler through the CLI. After $t_{total}$ has been computed, the upper bound of the loop is calculated according to \eqref{eq:t_validator}. 
% remove -> refer to equation in previous chapter
\begin{equation}
t = \lceil t_{total} / \#validators \rceil
\end{equation}

\texttt{generate\_forward} can be generated from a simple template with the id as the only variant and only consists of  an access to the storage and the emission of an internal operation.

\subsubsection{Output}
Besides the target code for the manager and assertion contracts, the compiler should return a piece of template code for the correct storage initialization of the manager. Since this code is passed to the origination operation, it must be in Michelson regardless of the compilation target language. The placeholders or accompanying info should be descriptive enough, s.t. the user can plug in the corresponding addresses after the origination of the parent and assertion contracts. For the example manager contract given in \secref{sec:modular}, the corresponding text output could be\\
\texttt{(Elt 0 <addr parent>; Elt 1 <addr A\_a>; Elt 2 <addr B\_a>)}.

\subsubsection{Direct compilation}\label{sec:direct}
\todo{verification by local node -> type check}
With the direct approach, the compiler generates Michelson code and hence code for a stack machine. Based on the simplicity of stack machines and the assumption that it does not need to generate optimal code, the compiler is expected to be simple \cite{cs5641}\cite{ferr_compiler}\cite{wiki:stack_machine}. As already indicate in the general part, the resulting contracts contain repeating structures and patterns for each input, which can be exploited by using template code during compilation. After compilation, the Michelson programs can be sanity-checked by type checking them using the provided Tezos libraries.

\paragraph{Assertion contracts}
\lstref{lst:contract_templ} shows a possible template for an assertion contract --- the placeholder \texttt{<?>} marks the locations where the individual code is inserted. Lines 3-5 prepare the stack, s.t. the parameter and storage are the two top elements. The lines below the placeholder for the body clean up the stack and leave it according to calling convention, i.e., the program returns an empty list of internal operations and the unit storage.
\lstinputlisting[label=lst:contract_templ, language=Michelson, caption=Template code for assertion contracts in Michelson, numbers=left]{listings/assertion_template.tz}

\paragraph{Manager contract}
The generation of the \texttt{parameter} and \texttt{storage} primitives is trivial --- the parameter type can be carried over directly from the parent contract, whereas the storage type is a constant (\texttt{(map int address)}). Based on the algorithm given previously, the code body is generated by nesting \texttt{IF\_LEFT} instructions recursively (cf. \lstref{lst:entrypoints}). The adapted algorithm is given in Algorithm \ref{alg:compile_manager_direct}.
\begin{algorithm}
\caption{Modification of algorithm \ref{alg:compile_manager} for direct compilation}\label{alg:compile_manager_direct}
	\begin{algorithmic}[0]
	\State ...
	\Function{compile}{parent parameter type}
	\If{ty = Or(l,r)} \Comment{Compile sum types to a branching}
	\State code\_left = "IF\_LEFT {" + compile(l) + "}"
	\State code\_right = "{" + compile(r) + "}"
	\State \Return code\_left + code\_right
	\EndIf
	\State ...
	\EndFunction
	\end{algorithmic}
\end{algorithm}

Owing to Michelson being a low-level language, the templates for the functions \texttt{generate\_loop} and \texttt{generate\_forward} go beyond the scope of this subsection. In order to provide an idea of how the generated code could look like, appendix \ref{apx:manager_michelson} contains two example contracts, one of which contains manager code for a non-empty and the other one for an empty assertion. The code was generated by the Liquidity compiler from two contracts written in Liquidity, as this provides a more compact and readable representation of the program logic. The corresponding source code can be found in \secref{sec:IR}.

\paragraph{Evaluation}
Compiling to Michelson directly is beneficial due of several reasons --- the pipeline stays streamlined and outputs the final contracts ready for origination. It does not introduce dependencies to external tools and their support for new protocol versions of Tezos. Besides, the new instructions only have to be added to Tezos' protocol code, whereas using external tools and languages would need to be patched as well in order to support them. \\
When it comes to verifying the compiler and integration testing of the pipeline, however, the low-level output causes more complexity than a high-level output. Due to the extend and restricted readability of non-trivial Michelson programs, their correctness is not always apparent in an instant. Tests can, on the one hand, compare the output with verified target contracts for a range of inputs, which requires a lot of Michelson code written by hand for each input. They, in turn, can also be prone to errors. \todo{remove?}On the other hand, the generated contracts can be tested as a black box using the test frameworks provided by Tezos, which use sandboxed networks to interact with contracts-under-test.

\subsubsection{Compiling to an IR}\label{sec:IR}
Instead of generating Michelson code directly, the compiler could also generate code in one of the high-level languages that provide a compiler with Michelson as a target. The output of the pipeline is then passed to this compiler in order to retrieve the final contracts. If the toolchain of the intermediate language doesn't provide a programmatic interface, the pipeline thus has to be extended with a separate step of procedure, as shown in \figref{fig:pipeline_liq}. This section demonstrates the implementation of this approach on the basis of Liquidity, hence the corresponding file extension \texttt{.liq}. \todo{provide Ligo as alternative -> could be included as package}
\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{figures/5-offline_tezos/pipeline_liq}
\caption{Pipeline with IR compilation and an external toolchain}
\label{fig:pipeline_liq}
\end{figure}

\paragraph{Assertion contracts}
\lstref{lst:contract_templ_liq} shows the equivalent of the assertion contract template given in \lstref{lst:contract_templ}. Since the Liquidity compiler takes care of the handling of the stack, it basically consists only of a storage type declaration, a function signature and the ``empty'' return type.
\lstinputlisting[label=lst:contract_templ_liq, language=Michelson, caption=Template code for assertion contracts in Liquidity, numbers=left]{listings/assertion_template.liq}

\paragraph{Manager contract}
Liquidity contracts are a list of OCaml-like functions that take a parameter and a storage. Instead of declaring the contract parameter type globally and nesting the code for each entrypoint, each entrypoint thus is declared as a separate function with a local parameter type. During compilation, the contract parameter type will then be composed according to the order of functions. 
Due to Algorithm \ref{alg:compile_manager} being left recursive, the output will automatically retain the correct order of entrypoints. In fact, the only adaptions necessary are the output of the two functions \texttt{generate\_loop} and \texttt{generate\_forward}, which should not only generate the code body, but return a coherent function. The function \texttt{entrypointA} in \lstref{lst:manager_templ_liq} is an example of such a function for an empty assertion. The non-empty version of the manager code can be found in appendix \ref{apx:apx:manager_liq}. As already noted, extracting a pure template for the code exceeds the scope of this section.
\lstinputlisting[label=lst:manager_templ_liq, language=Michelson, caption=Manager code in Liquidity for an entrypoint with an empty assertion, numbers=left]{listings/manager_forward.liq}

\paragraph{Evaluation}\todo{rename?}
This approach takes advantage of the usually more sophisticated high-level language compiler; they may provide options for optimizing generated code and are already thoroughly tested \cite{liq_repo}\cite{ligo_repo}. In contract to low-level code, the output of the pipeline compiler is easier to be verified, as it can be read like, for instance, an OCaml program. Providing target contracts for integration tests can be done much more efficiently and more reliable. \\
Besides the amenities of these compilers, this approach has a crucial disadvantage of introducing an external dependency. The compiler of the respective language has to be extended with the additional Michelson instructions as well, requiring a fork of the project. Furthermore one has to rely on continuous support of new protocol versions in the root project and integrate the updates into the own codebase. All things considered, the effort for extending both Michelson and the compiler of the IR, followed by a continuous effort of version management, may outweigh the benefits of this approach.

When returning to the orchestration approach of generating a monolithic contract, using Liquidity as an IR would allow to decompile the parent contract using the decompiler of its toolchain. Since the Liquidity compiler handles the composition of the contract parameter type and the code nesting, the assertion function simply have to be appended to the parent code, which significantly decreases the compilation complexity.

>>>>

\subsubsection{High-level languages}\label{sec:languages}
Michelson is a compilation target for various high-level languages that provide a more user-friendly and intuitive way of writing Smart Contracts. Additionally, some of them come with development environments, testing or verification tools. The following list introduces three prominent languages compiling to Michelson:
\begin{description}
\item[\textbf{Liquidity}] \cite{liquidity} is a language with an OCaml-like syntax, which allows using local variables instead of stack manipulations. Its module system can be used to write reusable contract code or libraries. Besides an optimizing compiler, the project also includes a decompiler to compile Michelson programs to Liquidity source code.
\item[\textbf{SmartPy}] \cite{smartpy} is a language available through a Python library and lets developers write contracts and tests using Python syntax and structures (such as classes). Its developer suite includes, i.a., a compiler, a simulation engine for testing contracts and an online editor.
\item [\textbf{Ligo}] \cite{ligo} offers multiple syntax flavours close to Pascal, OCaml or Reason. Besides compilation, the toolchain allows to invoke or evaluate the generated code locally. The project provides Ligo as OCaml packages, which can be integrated into other projects.
\end{description}

@@

Even though the monolithic approach is expected to be more efficient in terms of cost and network load, the following sections are based on the modular approach as the proposed architecture, due to its simplicity and flexibility.

@@

\subsection{Compiler}\label{sec:compiler}
The compiler finally translates the transformed and type checked assertion ASTs to Michelson contracts according to the modular orchestration scheme. Assertion and parent contract are thus assembled as depicted in \figref{fig:modular_assembly}. The compilation stage was not implemented within the scope of this thesis, but the following subsections describe the compilation procedures for generating the assertion and controller contracts. Furthermore, this section discusses the possibility to compile to a high-level language as an intermediate representation.

With Michelson as the target language, the compiler has to generate code for a stack machine. After compilation, the Michelson programs can be validated for correctness by the Tezos node. This will intercept any assertions containing type errors. After sanity-checking the programs, the compiler returns the target code for the controller and assertion contracts. As an additional output, it should return a piece of template code for the correct storage initialization of the controller, where the user can fill in the blockchain addresses of the other modules.

\subsubsection{Compilation of the assertion contracts}
The assertion ASTs are independent and can thus be compiled in isolation to separate Michelson contracts. Because they only have a single entrypoint, the parameter type corresponds to the raw parameter type of the parent entrypoint. If the parameter pattern in the assertion does not contain any wildcards, the parameter type can be derived directly from the pattern. Otherwise, the compiler has to infer the type from the parent contract.

If the pattern excludes any special values of the data type (e.g., \texttt{cons head tail} excludes empty lists), the generated code should reflect this. Excluded values should never cause the assertion to fail. In the example of \texttt{cons}, the generated code should thus contain the branching shown in \lstref{lst:cons}. The branch in line 1 is executed if the list contains at least one element, and should contain the assertion code. If the list is empty, the branch in line 2 should simply return successfully. 
\begin{lstlisting}[language=Michelson, label=lst:cons, caption=Exclusion of empty lists in the assertion]
IF_CONS ( <assertion body> )
        ( <return ([], ())>)
\end{lstlisting}
% Translation from quantifiers to random generators!

The generated code of the assertion body and the parameter type can be plugged into a template, which contains the necessary boilerplate code for a contract. Such a template is shown in \lstref{lst:contract_templ} --- the placeholders \texttt{<?>} mark the locations where the individual code is inserted. Lines 3-5 prepare the stack, s.t. the parameter and storage are the two top elements. The lines below the placeholder for the body clean up the stack and leave it according to calling convention, i.e., the program returns an empty list of internal operations and the unit storage.
\lstinputlisting[label=lst:contract_templ, language=Michelson, caption=Template code for assertion contracts in Michelson, numbers=left]{listings/assertion_template.tz}

\subsubsection{Compilation of the controller contract}
The generation of the \texttt{parameter} and \texttt{storage} primitives is trivial. Given that the controller acts as a proxy, the parameter type can be copied from the parent contract. Assuming that the proposal for the storage type from \lstref{lst:manager_modular} is adopted, the storage type is always the same, i.e., \texttt{(map int address)}. The code body is generated recursively based on the parameter type. It consists of two types of entrypoints, which either act as a direct gateway to the parent contract, or trigger an assertion check before invoking the parent.

\paragraph{Gateway entrypoints}
The code of gateway entrypoints only has to execute two steps; first, the address of the parent contract is read from the storage. It then creates an internal operation to this address, passing along the original parameter and funds. Assuming the parent address is always stored under the same key, e.g. with id 0, such gateways can be inserted as fixed boilerplate code.

\paragraph{Assertion entrypoints}
If the entrypoint should trigger an assertion, the compiler has to generated code executing the following steps:
\begin{enumerate}
\itemsep-0.5em
\item Read the addresses from the assertion and parent contract from storage
\item Calculate $t$
\item Compile a list of $t$ internal operations invoking the assertion with the raw parameter
\item Create an internal operation calling the parent contract, passing the original parameter and funds
\item Return a list of all operations and the unchanged storage
\end{enumerate}
\secref{sec:prob_threshold} stated a formula to determine the total number $t_{total}$ of required test runs for a given probability threshold $c$ and the size of the search space $n$. From $t_{total}$, the number of test runs $t$ per validator can be calculated according to \eqref{eq:t_validator}. Since $n$ can only be obtained at runtime, the compiler needs to derive a formula to determine it from the ranges of the random generators. \secref{sec:sample_space} already gave some examples of this. 
The threshold $c$, on the other hand, must be given as parameter to the compiler. Therefore, the command line interface of the toolchain should provide an option to set the threshold.

\paragraph{Algorithm}
Algorithm \ref{alg:compile_manager} shows the outline of a simplified algorithm, which builds the code body of the controller. It recursively iterates over the parameter type and builds the body by nesting \texttt{IF\_LEFT} instructions (cf. \lstref{lst:entrypoints}). Within the entrypoint branches, the compiler either generates gateway or assertion triggering code, based on the mapping given by the linker.

\begin{algorithm}
\caption{Simplified recursive algorithm for building the controller}\label{alg:compile_manager}
	\begin{algorithmic}[1]
	\State global storage = \{0: <parent>\}
	\State global i = 1
	\Function{compile\_controller}{parameter type}
	\If{ty = Or(l,r)}
	\State code\_left = "IF\_LEFT {" + compile(l) + "}"
	\State code\_right = "{" + compile(r) + "}"
	\State \Return code\_left + code\_right
	\Else
	\If{entrypoint has assertion}
	\State code = generate\_assertion\_trigger(parameter type, i, assertion)
	\State storage.add(i++, entrypoint tag)
	\Else
	\State code = generate\_gateway()
	\EndIf
	\State \Return code
	\EndIf
	\EndFunction
	\end{algorithmic}
\end{algorithm}

During this recursion, the compiler also builds the address map for the storage, such that the entrypoints can look up the address of the assertion and parent contracts during runtime. To this end, it keeps the counter \texttt{i} to create unique IDs for each entrypoint with an assertion --- this ID is then hard-wired into the storage access within this entrypoint. The address of the parent contract is always stored under ID 0.

After compilation, this map is translated to a template for a Michelson value and returned to the user together with the contract code. Because the addresses are not known at the time of compilation, the values should be descriptive place-holders, s.t. the user can replace them with the corresponding addresses. The completed Michelson value should then be passed to the origination as the initial storage. As an example, the compilation of the controller contract for \texttt{Example.tz} and \texttt{Example.tza} in Listings \ref{lst:linker_parent} and \ref{lst:linker_assertion} could return the following template:\\ \texttt{(Elt 0 <addr parent>; Elt 1 <addr A\_a>; Elt 2 <addr B\_a>)}.

In order to provide a detailed example, appendix \ref{apx:controller_michelson} contains the full Michelson code of a controller contract with one gateway and one assertion entrypoint. The code was generated by the Liquidity compiler from the source given in appendix \ref{apx:controller_liq}, which provides a more compact and readable representation of the program logic.

\subsubsection{Compiling to an intermediate representation}\label{sec:IR}
Instead of generating Michelson code directly, the compiler could also generate code in one of the high-level languages listed in \secref{sec:languages}. They all provide compilers targeting Michelson, and can thus serve as an intermediate representation (IR) of the contract code.

It depends on the language, how the compiler of the IR can be integrated into the toolchain. The Ligo project builds their compiler into a \texttt{dune}\footnote{dune \cite{dune} is a build system for OCaml} library, which can be imported directly into the code base of the toolchain. This keeps the pipeline coherent and would allow to output Michelson contracts without any intermediate steps. If the compiler cannot be embedded directly into the toolchain (as is the case with Liquidity), the output of the would have to be compiled manually to Michelson in a separate step, as depicted in \figref{fig:pipeline_liq}.

\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{figures/5-offline_tezos/pipeline_liq}
\caption{Pipeline with IR compilation and an external toolchain}
\label{fig:pipeline_liq}
\end{figure}

\paragraph{Comparison to direct compilation}
This approach takes advantage of the usually more sophisticated high-level language compiler; they may provide options for optimizing generated code and are already thoroughly tested. In contrast to large and obscure Michelson programs, the output is more comprehensible and thus easier to test and verify.

The crucial disadvantage of the IR approach is the introduction of an external dependency. The compiler of the respective language has to be extended with the \texttt{random} instruction as well, which requires forking the project. Furthermore, one has to rely on continuous support of new protocol versions in the root project and integrate the updates into the own code base regularly. To avoid such dependencies and increased effort in version management, this approach is not pursued in the development of the toolchain.

However, a realistic use-case for using Liquidity as an IR is the implementation of the monolithic orchestration scheme. In Liquidity, each entrypoint is declared as a function expecting a parameter and the current storage. During compilation, the contract parameter type and code branching is composed according to the order of functions. For instance, the entrypoint declarations in \lstref{lst:liq_eps} are compiled to the contract parameter type \texttt{(or (int \%A) (int \%B))} and two branches in the code body.  
\begin{lstlisting}[numbers=none, label=lst:liq_eps, caption=Entrypoint declarations in Liquidity]
let%entry A (a : int) storage = ...
let%entry B (b : int) storage = ...
\end{lstlisting}
As the Liquidity project provides a decompiler to translate Michelson contracts into Liquidity programs, the compiled assertion functions can simply be appended to the decompiled parent contract. This outsources most of the compilation complexity to the Liquidity compiler.

### 2021-04-25 assertions/doc/thesis-bernhardt/chapters/appendix.tex
Revision compiler

\chapter{Manager contract code in Liquidity}\label{apx:manager_liq}
\lstinputlisting[label=lst:manager_templ_liq, language=Michelson, caption=Manager code in Liquidity for an entrypoint with a non-empty assertion, numbers=left]{listings/manager_loop.liq}

\chapter{Manager contract code in Michelson}\label{apx:manager_michelson}
\lstinputlisting[basicstyle=\linespread{1.0}\fontfamily{lmr}\selectfont\small,
				 backgroundcolor=\color{cverbbg},
				 linewidth=14cm,
				 xleftmargin=0.5cm,
				 frame=lr,
				 framesep=8pt,
				 framerule=0pt,
				 captionpos=b,
				 numbers=none,
				 language=Michelson,
				 label=lst:manager_michelson_loop,
				 caption=Manager code in Michelson for an entrypoint with a non-empty assertion (wrapped in contract boilerplate code)]{listings/manager_loop.tz}

\lstinputlisting[basicstyle=\linespread{1.0}\fontfamily{lmr}\selectfont\small,
				 backgroundcolor=\color{cverbbg},
				 linewidth=14cm,
				 xleftmargin=0.5cm,
				 frame=lr,
				 framesep=8pt,
				 framerule=0pt,
				 captionpos=b,
				 numbers=none,
				 language=Michelson,
				 label=lst:manager_michelson_forward,
				 caption=Manager code in Michelson for an entrypoint with an empty assertion (wrapped in contract boilerplate code)]{listings/manager_forward.tz}

>>>>

\chapter{Controller contract code in Liquidity}\label{apx:controller_liq}
\lstinputlisting[label=lst:manager_templ_liq, language=Michelson, caption=Controller code in Liquidity]{listings/controller.liq}

\chapter{Controller contract code in Michelson}\label{apx:controller_michelson}
\lstinputlisting[basicstyle=\linespread{1.0}\fontfamily{lmr}\selectfont\small,
				 backgroundcolor=\color{cverbbg},
				 linewidth=14cm,
				 xleftmargin=0.5cm,
				 frame=lr,
				 framesep=8pt,
				 framerule=0pt,
				 captionpos=b,
				 numbers=none,
				 language=Michelson,
				 label=lst:controller_michelson,
				 caption=Controller code in Michelson]{listings/controller.tz}
